Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}

### 2018-02-14 bn.tex
Minor typos

A Bayesian network $B=(G,\theta)$ defines a probability distribution for
$X$. The component $G$ defines the structure of the model as a
directed acyclic graph (DAG) that has exactly one node for each component of
$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
the nodes from which there are a directed edges to the variable
$X_{i}$.

@@

Score-based Bayesian learning of Bayesian network structures evaluates the goodness of different structures $G$ using their
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyperparameters for the model parameters $\theta$, and the $D$ is a
collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
collected to a $N\times n$ design matrix. We use the notation $D_i$ to
denote the $i^\text{th}$ column of the data matrix and notation $D_V$ to denote
the columns that correspond to the variable subset $V$. We also write
$D_{i,G_i}$ for $D_{\{i\}\cup G_i}$ and denote the entries of the
column $i$ on the rows on which the parents $G_i$ contain the value
configuration number by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.

It is common to assume the uniform prior for structures, in which case
the objective function for structure learning is reduced to the
marginal likelihood $P(D|G,\alpha)$.  If the model parameters
$\theta_{ij}$ are further assumed to be independently Dirichlet
distributed only depending on $i$ and $G_{i}$ and the data $D$ is
assumed to have no missing values, the marginal likelihood can be
decomposed as
\begin{eqnarray}
\label{eqn:bayesmix}
\lefteqn{P(D|G,\alpha)}\nonumber\\
&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}P(D_{i,G_i=j};\alpha)\nonumber\\
&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}\int P(D_{i,G_i=j}|\theta_{ij})P(\theta_{ij};\alpha) d\theta_{ij}.
\end{eqnarray}
In coding terms this means that each data column $D_i$ is first
partitioned based on the values in columns $G_i$, and each part is
then coded using a Bayesian mixture that can be expressed in a closed
form~\cite{Bunt91, Heck95}.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the empirical entropy $H_N(X_i | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates \textit{regularity} in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent. In Section \ref{sec:regularity}, we will give the formal definition of regularity and state that qNML is regular. In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before. The detailed proofs can be found in Appendix B in the Supplementary Material.

A natural solution to avoid parameter sensitivity of BDeu would be to
use a normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
maximizes
\begin{equation}
P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $\hat\theta$ denotes the (easy to find) maximum likelihood
parameters and the sum in the denominator goes over all the possible
$N\times n$ data matrices. This information-theoretic NML criterion
can be justified from the minimum description length point of view,
\cite{Riss78,Grun07}. It has been shown to be robust with respect to
different data generating mechanisms where a good choice of prior
is challenging, see~\cite{eggeling2014robust,maatta16}. While it is
easy to see that the NML criterion satisfies the requirement of giving
equal scores to equal structures, the normalizing constant renders the
computation infeasible.

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML-code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

>>>>

A Bayesian network $B=(G,\theta)$ defines a probability distribution for
$X$. The component $G$ defines the structure of the model as a
directed acyclic graph (DAG) that has exactly one node for each component of
$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
the nodes from which there is a directed edge to the variable
$X_{i}$.

@@

Score-based Bayesian learning of Bayesian network structures evaluates the goodness of different structures $G$ using their
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyperparameters for the model parameters $\theta$, and $D$ is a
collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
collected to a $N\times n$ design matrix. We use the notation $D_i$ to
denote the $i^\text{th}$ column of the data matrix and the notation $D_V$ to denote
the columns that correspond to the variable subset $V$. We also write
$D_{i,G_i}$ for $D_{\{i\}\cup G_i}$ and denote the entries of the
column $i$ on the rows on which the parents $G_i$ contain the value
configuration number by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.

It is common to assume the uniform prior for structures, in which case
the objective function for structure learning is reduced to the
marginal likelihood $P(D|G,\alpha)$.  If the model parameters
$\theta_{ij}$ are further assumed to be independently Dirichlet
distributed only depending on $i$ and $G_{i}$ and the data $D$ is
assumed to have no missing values, the marginal likelihood can be
decomposed as
\begin{eqnarray}
\label{eqn:bayesmix}
\lefteqn{P(D|G,\alpha)}\nonumber\\
&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}P(D_{i,G_i=j};\alpha)\nonumber\\
&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}\int P(D_{i,G_i=j}|\theta_{ij})P(\theta_{ij};\alpha) d\theta_{ij}.
\end{eqnarray}
In coding terms this means that each data column $D_i$ is first
partitioned based on the values in columns $G_i$, and each part is
then coded using a Bayesian mixture that can be expressed in closed
form~\cite{Bunt91, Heck95}.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has the tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$ the values of $X_i$ are completely determined by variables
in set $Z$, so that the empirical entropy $H_N(X_i | Z)$ is zero. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behavior violates \textit{regularity} in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent. In Section \ref{sec:regularity}, we will give the formal definition of regularity and state that qNML is regular. In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before. The detailed proofs can be found in Appendix B in the Supplementary Material.

A natural solution to avoid parameter sensitivity of BDeu would be to
use a normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
maximizes
\begin{equation}
P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $\hat\theta$ denotes the (easy to find) maximum likelihood
parameters and the sum in the denominator goes over all the possible
$N\times n$ data matrices. This information-theoretic NML criterion
can be justified from the minimum description length point of view
\cite{Riss78,Grun07}. It has been shown to be robust with respect to
different data generating mechanisms where a good choice of prior
is challenging, see~\cite{eggeling2014robust,maatta16}. While it is
easy to see that the NML criterion satisfies the requirement of giving
equal scores to equal structures, the normalizing constant renders the
computation infeasible.

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

### 2018-02-14 experiments.tex
mention the SHD ranking in case of ties
Minor typos

In our first experiment, we took five widely used benchmark Bayesian networks\footnote{Bayesian Network Repository: \url{http://www.bnlearn.com/bnrepository/}}, sampled data from them, and tried to learn the generating structure with the different scoring functions using various sample sizes. We used the following networks: Asia ($n=5$, 8 arcs), Cancer ($n=5$, 4 arcs), Earthquake ($n=5$, 4 arcs), Sachs ($n=11$, 17 arcs) and Survey ($n=6$, 6 arcs). These networks were picked in order to use the dynamic programming based exact structure learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20. We measured the quality of the learned structures using structural Hamming distance (SHD, \cite{Tsamardinos2006}).

Figure \ref{fig:all_shd} shows SHDs for all the scoring criteria for each network. Sample sizes range from $10$ to $10000$ and the shown results are averages computed from $1000$ repetitions. None of the scores dominates in all settings considered. BIC fares well when the sample size is small, as it tends to produce a nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse. qNML obtains strong results in 'Earthquake' and 'Asia' networks, being the best or the second best with all the sample sizes considered.

@@

Figure \ref{fig:shd_ranks} summarizes the SHD results in all networks by showing the average rank for each score. The ranking was done by giving the score with the lowest SHD rank $1$ and the worst one rank $4$. The shown results are averages computed from $5000$ values ($5$ networks, $1000$ repetitions). From this, we can see that qNML never has the worst average ranking, and it has the best ranking with sample sizes greater than $300$. This suggests that qNML is overall a safe choice in structure learning, especially with moderate and large sample sizes.

@@

When predicting $P(d_{test}|D_{train},S,\theta)$,
with structures $S$ learned by the BDeu score, we used the
Bayesian predictive parameter values (BPP) $\theta_{ijk} \propto
N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
hyperparameter-free, for structures learned by the other model
selection criteria, we used the sequential predictive NML (sNML)
parametrization $\theta_{ijk}\propto e(N_{ijk})(N_{ijk}+1)$, where
$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.

>>>>

In our first experiment, we took five widely used benchmark Bayesian networks\footnote{Bayesian Network Repository: \url{http://www.bnlearn.com/bnrepository/}}, sampled data from them, and tried to learn the generating structure with the different scoring functions using various sample sizes. We used the following networks: Asia ($n=5$, 8 arcs), Cancer ($n=5$, 4 arcs), Earthquake ($n=5$, 4 arcs), Sachs ($n=11$, 17 arcs) and Survey ($n=6$, 6 arcs). These networks were picked in order to use the dynamic programming based exact structure learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20. We measured the quality of the learned structures using structural Hamming distance (SHD) \cite{Tsamardinos2006}.

Figure \ref{fig:all_shd} shows SHDs for all the scoring criteria for each network. Sample sizes range from $10$ to $10000$ and the shown results are averages computed from $1000$ repetitions. None of the scores dominates in all settings considered. BIC fares well when the sample size is small as it tends to produce a nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse. qNML obtains strong results in the Earthquake and Asia networks, being the best or the second best with all the sample sizes considered.

@@

Figure \ref{fig:shd_ranks} summarizes the SHD results in all networks by showing the average rank for each score. The ranking was done by giving the score with the lowest SHD rank $1$ and the worst one rank $4$. In case of ties, the methods with the same SHD were given the same rank. The shown results are averages computed from $5000$ values ($5$ networks, $1000$ repetitions). From this, we can see that qNML never has the worst average ranking, and it has the best ranking with sample sizes greater than $300$. This suggests that qNML is overall a safe choice in structure learning, especially with moderate and large sample sizes.

@@

When predicting $P(d_{test}|D_{train},S,\theta)$
with structures $S$ learned by the BDeu score, we used the
Bayesian predictive parameter values (BPP) $\theta_{ijk} \propto
N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
hyperparameter-free, for structures learned by the other model
selection criteria, we used the sequential predictive NML (sNML)
parametrization $\theta_{ijk}\propto e(N_{ijk})(N_{ijk}+1)$, where
$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.

### 2018-02-14 intro.tex
Minor typos

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information criterion
(AIC)~\cite{Akai73} and the Bayesian information criterion
(BIC)~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for AIC, and BIC appears to
require large sample sizes in order to identify appropriate
structures~\cite{cosco.pgm08a,Liu2012}. Traditionally, the most
popular criterion has been the Bayesian marginal
likelihood~\cite{Heck95b} and its BDeu variant (see
section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} have
found this criterion to be very sensitive to hyperparameters and to
yield undesirably complex models for small sample sizes.

The information-theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be a potential
candidate for a good criterion, but its exact calculation is likely to
be prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike fNML,
qNML is \textit{score equivalent}, i.e., it gives equal scores to
structures that encode the same independence and dependence
statements. Like other common model selection criteria, qNML is also
consistent.

We next briefly introduce Bayesian networks and review the BDeu and
fNML criteria and then introduce the qNML criterion.  We also
summarize the results for 20 data sets to back up our claim that qNML
yields parsimonious models with good predictive capabilities. The
experiments with artificial data generated from real world Bayesian networks demonstrate the capability of our score to quickly learn a structure close to the generating one.

>>>>

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information criterion
(AIC)~\cite{Akai73} and the Bayesian information criterion
(BIC)~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for AIC, and BIC appears to
require large sample sizes in order to identify appropriate
structures~\cite{cosco.pgm08a,Liu2012}. Traditionally, the most
popular criterion has been the Bayesian marginal
likelihood~\cite{Heck95b} and its BDeu variant (see
Section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} have
found this criterion to be very sensitive to hyperparameters and to
yield undesirably complex models for small sample sizes.

The information-theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be a potential
candidate for a good criterion, but its exact calculation is likely to
be prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter-free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike fNML,
qNML is \textit{score equivalent}, i.e., it gives equal scores to
structures that encode the same independence and dependence
statements. Like other common model selection criteria, qNML is also
consistent.

We next briefly introduce Bayesian networks and review the BDeu and
fNML criteria and then introduce the qNML criterion.  We also
summarize the results for 20 data sets to back up our claim that qNML
yields parsimonious models with good predictive capabilities. The
experiments with artificial data generated from real-world Bayesian networks demonstrate the capability of our score to quickly learn a structure close to the generating one.
